Stochastic Gradient Descent

### 2018-03-21 hw3/hw3.tex
finished final presentation

We know that $\mathbf{U}^T\mathbf{U}=\mathbf{I}$ and $\mathbf{D}^T\mathbf{D}=\mathbf{D}^2$ and so can
simplify this to:
$$
= (\mathbf{VD}^2\mathbf{V}^T + \lambda \mathbf{I})^{-1} \mathbf{VDU}^T\mathbf{y}
$$

We substitute in $\mathbf{VV}^T$ for $\mathbf{I}$:

@@

And make use of the fact that $\mathbf{V}$ is orthonormal:
$$
= \mathbf{V}(\mathbf{D}^2 + \lambda)^{-1}\mathbf{DU}^T\mathbf{y}
$$

@@

In the inefficient method, we recompute the SVD every
time we update $\lambda$. SVD has complexity on the order of
$O(nd^2)$ (with $n$ corresponding to the number of rows of the feature matrix and 
$d$ corresponding to the number of columns). We then multiply $\mathbf{VD}_\lambda \mathbf{U}^T\mathbf{y}$.
A matrix-matrix product $C=AB$, where $A \in \mathbb{R}^{m \times n}$ and $B \in
\mathbb{R}^{n \times p}$, costs $2mnp$ flops. In our case, $\mathbf{U}^T \in
\mathbb{R}^{d \times n}$ and $\mathbf{Y} \in \mathbb{R}^{n \times 1}$, and so
multiplying the two costs $2 \times n \times d \times 1 = 2nd$ flops and results
in a $d \times 1$ vector. Multiplying $\mathbf{D}_\lambda \in \mathbb{R}^{d
\times d}$ and this $d \times 1$ vector costs $d$ flops (assuming we take advantage of the
diagonal structure of $\mathbf{D}_\lambda$) and results in a $d \times 1$ vector. 
Multiplying $\mathbf{V} \in \mathbb{R}^{d \times d}$ by this $d \times 1$ vector costs 
$2d^2$ flops. Adding together all these steps, we have $nd^2 + 2nd + d +
2d^2$ flops. Dropping all constant coefficients and only considering the
highest-order polynomial, we conclude that the inefficient method costs
$O(nd^2)$ flops.

>>>>

$\mathbf{U}$ is orthogonal ($\mathbf{U}^T\mathbf{U}=\mathbf{I}$) and
$\mathbf{D}$ is diagonal ($\mathbf{D}^T = \mathbf{D}$):
$$
= (\mathbf{VD}^2\mathbf{V}^T + \lambda \mathbf{I})^{-1} \mathbf{VDU}^T\mathbf{y}
$$

Because $\mathbf{V}$ is orthonormal, i.e. $\mathbf{V}^T = \mathbf{V}^{-1}$, we substitute in $\mathbf{VV}^T$ for $\mathbf{I}$:

@@

And again make use of the fact that $\mathbf{V}$ is orthonormal, this time to
cancel some terms:
$$
= \mathbf{V}(\mathbf{D}^2 + \lambda)^{-1}\mathbf{DU}^T\mathbf{y}
$$

@@

In the inefficient method, we recompute the SVD every
time we update $\lambda$. SVD has complexity on the order of
$O(nd^2)$ (with $n$ corresponding to the number of rows of the feature matrix and 
$d$ corresponding to the number of columns). We then multiply
$\mathbf{VD}_\lambda \mathbf{U}^T\mathbf{y}$. We exploit the associative
property of matrices and proceed right to left for maximum efficiency, 
i.e. $\mathbf{V}(\mathbf{D}_\lambda(\mathbf{U}^T\mathbf{y}))$.
A matrix-matrix product $C=AB$, where $A \in \mathbb{R}^{m \times n}$ and $B \in
\mathbb{R}^{n \times p}$, costs $2mnp$ flops. In our case, $\mathbf{U}^T \in
\mathbb{R}^{d \times n}$ and $\mathbf{Y} \in \mathbb{R}^{n \times 1}$, and so
multiplying the two costs $2 \times n \times d \times 1 = 2nd$ flops and results
in a $d \times 1$ vector. Multiplying $\mathbf{D}_\lambda \in \mathbb{R}^{d
\times d}$ and this $d \times 1$ vector costs $d$ flops (assuming we take advantage of the
diagonal structure of $\mathbf{D}_\lambda$) and results in a $d \times 1$ vector. 
Multiplying $\mathbf{V} \in \mathbb{R}^{d \times d}$ by this $d \times 1$ vector costs 
$2d^2$ flops. Adding together all these steps, we have $nd^2 + 2nd + d +
2d^2$ flops. Dropping all constant coefficients and only considering the
highest-order polynomial, we conclude that the inefficient method costs
$O(nd^2)$ flops.

### 2018-03-21 sgd/sgd.tex
finished final presentation

\documentclass{beamer}
\usetheme{Madrid}

@@

\usepackage{graphicx}
\graphicspath{ {img/} }

@@

	In \textbf{parametric} statistics, we often make assumptions that make this optimization ``nice'':
	\begin{itemize}
		\item \small Example: in OLS, we do not need to check every possible value
			of $\hat{\beta}$ to see if it minimizes the loss
			function, we (typically) can just evaluate
			$(X'X)^{-1}X'Y$ \\~\\
	\end{itemize}
	\smallskip

@@

\begin{frame}
	\frametitle{Other times, optimization is not so easy}
	Suppose that we have a typical supervised classification problem:
	\begin{itemize}
		\item \small Non-parametric: no assumptions about distribution of data
		\item Feature vector $\mathbf{X}_i$, label $Y_i$
		\item Want to find best prediction function $f^*$ from class $\mathcal{F}$
		\item $f$ has parameter vector $\bm{\theta}$
		\item Optimization: pick values for $\bm{\theta}$ that minimize
			empirical risk according to some
			loss function $L$
		\item Assume $L$ is convex (if not, problem is \textit{much}
			harder)\\~\\
	\end{itemize}

@@

\begin{frame}
	\frametitle{Gradient descent is a numerical approach to optimization}
	Intuition: gradient is multidimensional version of 1st derivative
	\begin{itemize}
		\item \small $\nabla f(\mathbf{X}) = (\frac{\partial f}{x_1},
			\frac{\partial f}{x_2}...)$
		\item So if the gradient is (approximately) 0 at some point, that point is
			(approximately) a stationary point 
		\item Since we know the loss function is convex, that stationary
			point must be the global minimum\\~\\
	\end{itemize}

	Give basic idea: take guess, step, evaluate, stop once below epsilon
\end{frame}

\begin{frame}
	\frametitle{Batch gradient descent, animated}
\end{frame}

\begin{frame}
	\frametitle{Batch gradient descent is computationally expensive}
	Elaborate
\end{frame}

\begin{frame}
	\frametitle{Stochastic gradient descent (SGD) is more efficient}
	Overview of how it works \\~\\
\end{frame}

\begin{frame}
	\frametitle{SGD, visualized}
\end{frame}

\begin{frame}
	\frametitle{Choice of hyper-parameters is important}
	Reminder of what they are \\~\\
	Step size from Bottou \\~\\
	Learning parameter
\end{frame}

\begin{frame}
	\frametitle{SGD has nice properties for high-dimensional
	data}
	Theoretical explanation \\~\\
	Empirical performance \\~\\

@@

\begin{frame}
	\frametitle{Applications of SGD}
	If a Silicon Valley press release uses any of the following phrases...
	\begin{itemize}
		\item \small ``Neural networks''
		\item ``Machine learning''
		\item ``AI'' 
	\end{itemize}

	...SGD probably is involved. Example: Google's \texttt{AlphaGo} program.

	\begin{figure}[b]
	\centering
	\includegraphics[scale=0.33]{go}
	\end{figure}

>>>>

\documentclass{beamer}
\usetheme{Madrid}
\usecolortheme{spruce}

@@

\usepackage{color}

\usepackage{graphicx}

\graphicspath{{./img/}}

@@

	In \textbf{parametric} statistics, we often make assumptions that make this optimization ``nice'':
	\begin{itemize}
		\item \small Example: in OLS, we do not need to try
			different values
			of $\hat{\beta}$ to see which minimizes the loss
			function, we (typically) can just evaluate
			$(X'X)^{-1}X'Y$ \\~\\
	\end{itemize}
	\smallskip

@@

\begin{frame}
	\frametitle{Other times, optimization is not so easy}
	Suppose that we have a typical supervised classification problem:
	\begin{itemize}
		\item \small Non-parametric: no assumptions about distribution of data
		\item Feature vector $\mathbf{X}_i$, label $Y_i$
		\item Want to find best prediction function $f_w^*$ from class $\mathcal{F}$
		\item Optimization: pick weights $\bm{w}$ that minimize
			empirical risk according to some convex
			loss function $L(f_w(x), y)$ \\~\\
	\end{itemize}

@@

\begin{frame}
	\frametitle{Gradient descent is an iterative search procedure}
	\begin{figure}[h]
	\centering
	\includegraphics[scale=0.33]{Sketch}
	\end{figure}
\end{frame}

\begin{frame}
	\frametitle{A more formal explanation}

	$\mathbf{w}_{t+1} = \mathbf{w}_t - \gamma \frac{1}{n}\sum_{i=1}^n
	\nabla_w L(f_w(\mathbf{X}_i), y_i)$ \\~\\

	Stop if $\mathbf{w}_t - \ldots \leq \epsilon$ \\~\\

	Step size $\gamma$ can vary over time \\~\\

	Theoretical guarantees on speed of convergence ($\rho :=$ size of error):
	\begin{itemize}
		\item Version presented here: $ - \log \rho \sim t$
		\item More optimized version: $ - \log \log \rho \sim t$ \\~\\
	\end{itemize}

	But, big difference between:
	\begin{itemize}
		\item Speed $:=$ number of iterations
		\item Speed $:=$ time (clock on wall)
	\end{itemize}
\end{frame}

\begin{frame}
	\frametitle{Batch gradient descent is computationally expensive}
	In ``plain'' (batch) gradient descent, for \textcolor{red}{every step}, 
	we have to evaluate the gradient at \textcolor{blue} {every observation}

	$$
	\textcolor{red}{\mathbf{w}_{t+1}} = \mathbf{w}_t - \gamma \frac{1}{n}
	\textcolor{blue}{\sum_{i=1}^n}
	\nabla_w L(f_w(\mathbf{X}_i), y_i)
	$$\\~\\

	This becomes computationally intractable as $n$ grows large \\~\\

	Knowing that the quality of our approximation gets linearly or
	quadratically better with $t$ is not comforting if each $t$ takes days to run
\end{frame}

\begin{frame}
	\frametitle{Stochastic gradient descent (SGD) takes less time}

	For each step, gradient is computed for a \textbf{single} randomly
	chosen observation $i$:

	$$
	{\mathbf{w}_{t+1}} = \mathbf{w}_t - \gamma \nabla_w L(f_w(\mathbf{X}_i), y_i)
	$$ \\~\\

	This simplification makes approximation much ``noisier'', and hence SGD
	requires more iterations \\~\\

	But, each iteration is faster and so SGD can reach a predefined
	level of risk or error in less time
\end{frame}

\begin{frame}
	\frametitle{SGD is particularly useful when $n$ is large and computation
	time is important}
	\begin{table}[t]
		\begin{tabular}{|l|c|c|}
			\hline
			& \textbf{GD} & \textbf{SGD} \\
			\textbf{Time to accuracy} $\mathbf{\rho}$ & $n \log
			\frac{1}{\rho}$ & $\frac{1}{\rho}$ \\
			\hline
		\end{tabular}
	\end{table}

	\begin{figure}
	\centering
	\includegraphics[scale=0.55]{comparison}
	\end{figure}

@@

\begin{frame}
	\frametitle{SGD is widely used in industry}
	If a Silicon Valley press release uses any of the following phrases...
	\begin{itemize}
		\item \small ``Neural networks''
		\item ``Machine learning''
		\item ``AI'' 
	\end{itemize}

	...SGD probably is involved. Example: Google's \textbf{AlphaGo} program.

	\begin{figure}[b]
	\centering
	\includegraphics[scale=0.25]{go}
	\end{figure}
